---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, echo=FALSE}
    # Set the numerical type of the output
    options(scipen = 5, digits = 2) 
```

## Loading and preprocessing the data
Let's load the data and take a quick look at its structure:
```{r, echo=TRUE}
    unzip("activity.zip")
    table <- read.csv("activity.csv", header=TRUE, na.strings="NA")
    str(table)
```

Now let's use the dplyr package in order to better manipulate
the time-type data:
```{r, echo=TRUE}
    # If dplyr is not installed please do: install.packages(dplyr)
    suppressMessages(library(dplyr))
    table <- mutate(table, date=as.Date(date))
```

## What is mean total number of steps taken per day?

Let us first calculate the total number of steps taken each day, ignoring the missing values (N.A):
```{r, echo=TRUE}
    steps_by_day <- na.omit(group_by(table, date))
    total_by_day <- summarize(steps_by_day, steps=sum(steps))
    head(total_by_day)
```
    
Next let's take a look at the histogram of the total number of steps by day:
    
```{r, echo=TRUE}
    hist(total_by_day$steps,breaks=10, col='yellow',
    main='Total number of steps per day', xlab='Number of steps')
```

And finally let's calculate the mean and the median values of the distribution:
```{r, echo=TRUE}
    mean_value <- mean(total_by_day$steps)
    median_value <- median(total_by_day$steps)
```

As for the total number of steps taken per day, the mean value is `r mean_value`, and the median is `r median_value`. Both numbers are virtually identical for this data set.

## What is the average daily activity pattern?
Let us start by estimating the mean number of steps of every 5-minute interval of the day, across all the differente days:
```{r, echo=TRUE}
    intervalos <- na.omit(group_by(table, interval))
    mean_interv <- summarize(intervalos, steps=mean(steps))
    head(mean_interv)
```

And now let's take a look ate the time-series plot:
```{r, echo=TRUE}    
    plot(mean_interv$interval, mean_interv$steps,
     main="Mean number of steps at each 5-minute interval",
     ylab="Number of steps", xlab="Time (in units of 5-minutes intervals)")
```
    
Finally, let's find the time interval at which the maximum number of steps occurs:
```{r, echo=TRUE}    
    index <- which.max(mean_interv$steps)
    timeofmax <- mean_interv$interval[index]
```

The 5-minute interval at which the average number of steps -across all the tabulated days- is maximum is the interval number: `r timeofmax`


## Imputing missing values

Let's calculate the total number of missing values:
```{r, echo=TRUE} 
    missingvals <- sum(is.na(table))
```

There are `r missingvals` missing values (NA) in this data set.


A good strategy to deal with the missing values may be replacing them by the average value at the same time interval -across the rest of the tabulated days-. So, let's create a vector of the full data set, look for all the missing values in it, and replace each of them by the mean value of the corresponding time interval:

```{r, echo=TRUE}
    new_values <- table$steps
    for (i in 1:length(new_values)){
        if (is.na(table$steps[i])){
            index <- which(mean_interv$interval==table$interval[i])
            new_values[i] <- mean_interv$steps[index]
        }
    }
```

And now let's create a new table, which is a copy of the original one, but replacing the NAs by the quoted averages; i.e., let's arrange the vector created in the former step to the original form:

```{r, echo=TRUE}
    new_table <- mutate(table, steps=new_values)
    head(new_table)
```

Next let's take a look at the histogram of the total number of steps by day:
    
```{r, echo=TRUE}
    steps_by_day <- na.omit(group_by(new_table, date))
    total_by_day <- summarize(steps_by_day, steps=sum(steps))
    hist(total_by_day$steps,breaks=10, col='yellow',
    main='Total number of steps per day', xlab='Number of steps')
```

And finally let's calculate again the mean and median values of the distribution of total number of steps taken each day:

```{r, echo=TRUE}
    mean_value <- mean(total_by_day$steps)
    median_value <- median(total_by_day$steps)
```
    
After replacing the missing data, the mean value of the total number of steps taken per day is `r mean_value`, and the median is `r median_value`. Now both numbers are exactly the same, and they coincide with the mean value obtained for the original data set.



## Are there differences in activity patterns between weekdays and weekends?

So, first of all we will create a new factor indicating the category (weekdays / weekends) each day belongs to:
```{r, echo=TRUE}
    new_table <- mutate(new_table, day=weekdays(date),     
    day=as.factor(weekdays(date)))
    levels(new_table$day) <- c("weekend", "weekday", "weekday", "weekend", "weekday",
    "weekday", "weekday")
    # I note that my system is taking the names of the days in portuguese, which make the order of the levels vary with respect to english
```

Finally, let's plot the time-series at weekdays and weekends to see how they compare to each other:
```{r, echo=TRUE}
    intervalos <- group_by(new_table, interval, day)
    mean_interv <- summarize(intervalos, steps=mean(steps))

    library(lattice)
    xyplot(mean_interv$steps ~ mean_interv$interval | mean_interv$day, 
    type = "l", xlab="Interval", ylab="Number of steps", layout=c(1,2))
```


